Black swans in ecological time series\\ Black swans in population dynamics\\ Heavy-tailed ecological time series\\ Black swans in ecological dynamics\\ Black swan ecological dynamics\\ \ldots

### 2014-11-25 text/ms.tex
Switch back bib style and bump version
A round of minor edits

Black swans are statistically improbable events that nonetheless occur---often
with profound consequences. While extremes in the physical environment, such
as monsoons and heat waves, are widely studied and increasing in magnitude and
frequency, it remains unclear the extent to which ecological populations
buffer or suffer from such extremes. Here, we develop a probability model to
estimate the degree of heavy-tailedness (presence of black swans) in
ecological process noise. We apply our model to \NPops~time series from around
the world across \NOrders~taxonomic orders and seven classes. We find strong
evidence of black swans, but they are rare, occurring in
\overallMinPerc--\overallMaxPerc\% of populations: most frequently for birds
(\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), and insects
(\InsectaRangePerc\%). Black swans were predominantly (\percBSDown \%)
downward events and were not explained by any life-history covariates, but
tended to be driven by external perturbations such as climate, severe winters,
predator or parasite cycles, and the combined effects of multiple factors.
Extreme events were more frequently detected for populations with longer time
series and lower levels of process noise; for shorter and noisier time series,
our simulations suggested black-swan dynamics are often not identifiable as
such. The presence of black swans in population dynamics highlights the
importance of developing robust conservation and management
strategies---particularly as the frequency and magnitude of climate extremes
increase over the next century.

@@

Black swans are unexpected extreme events with potentially dramatic
consequences \citep{taleb2007,sornette2009}. One of the most striking black
swans in ecology is the possible asteroid collision marking the end-Cretaceous
mass extinction 65 million years ago \citep{alvarez1980,harnik2012}. Today,
climate extremes, in concert with shifts in mean temperature, are expected to
cause the greatest ecological and societal damage \citep{ipcc2012}. But, while
extremes in the physical environment such as wave height, storm severity, and
temperature are frequent events \citep{gaines1993,katz2005}, it remains
unclear the extent to which ecological systems buffer or suffer from black
swans \citep{nunez2012}.

There is compelling anecdotal evidence for ecological black swans, but
systematic evidence across taxa has been elusive. A survey of ecologists
indicated that surprising outcomes of field experiments are far more common
than we assume \citep{doak2008}, and events such as the global invasion of
Argentine ants and the mutation of viruses to infect new hosts could be
considered black swans \citep{nunez2012}. In fact, anecdotes of population
catastrophes are numerous and catastrophes may be the most important element
affecting population persistence \citep{mangel1994}. For marine mammal
populations, we have compelling evidence of catastrophes \citep{gerber2001,
  ward2007}, and recently, time-series of marine microbe abundance
\citep{segura2013} and time to extinction for experimental waterflea
populations \citep{drake2014} have been found to have heavier tailed
distributions than the normal. Despite these anecdotes, as far as we can tell
there have been two systematic surveys for ecological black swans: one with
North American breeding birds and the other with the Global Population
Dynamics Database, which uncovered little clear evidence for them
\citep{keitt1998,allen2001,halley2002}. Nevertheless there are methodological
challenges to their detection \citep{allen2001,ward2007}.

@@

Our main analysis focuses on the commonly applied Gompertz population dynamics
model \citep[e.g.][]{knape2012,dennis2014,connors2014}. The Gompertz model
represents population growth as a linear function in $\log$ space. If we let
$x_t$ represent the $\log$ abundance ($N$) at time $t$, we can represent the
Gompertz model as:
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma).
\end{align*}
The growth parameter $\lambda$ represents the expected growth rate if $N_t =
1$. The model is density independent if $b = 1$, maximally density dependent if
$b = 0$, and inversely density dependent if $b < 0$. Usually, the process noise
$\epsilon_t$ is modelled as normally distributed, but in our paper we assume it
is drawn from a t distribution with scale parameter $\sigma$ and degrees of
freedom $\nu$. In previous analyses of the GPDD, the Gompertz was most often
identified as the most parsimonious population model that can be fit to these
data \citep{brook2006}.

@@

We chose weakly informative priors to incorporate our understanding of
plausible population dynamics (\citeauthor{gelman2014} \citeyear{gelman2014};
Figs~\ref{fig:didactic}f--h, \ref{fig:priors}). For $\nu$, we chose an
exponential prior with rate parameter of $0.01$ truncated at values above
two---a slightly less informative prior than suggested by
\citet{fernandez1998}. This prior gives only a \basePriorProbHeavy \%
probability that $\nu < 10$ but constrains the sampling sufficiently to avoid
wandering off towards infinity. In any case, for $\nu > 20$ the t distribution
is almost indistinguishable from the normal distribution
(Fig.~\ref{fig:didactic}). Based on the shape of the t distribution, we chose
the probability of $\nu < 10$, Pr($\nu < 10$), to define the probability of
heavy-tailed dynamics. When categorizing a population as heavy or normal tailed,
we used a threshold of 0.5 probability.

We fit alternative population models to test if four key phenomena
systematically changed our conclusions. Autocorrelation has been suggested as
a reason for increased observed variability of abundance time series through
time, which could create apparent heavy tails \citep{inchausti2002};
therefore, we fit a model that included serial correlation in the residuals.
Additionally, previous work has modelled abundance or growth rates without
accounting for density dependence \citep{halley2002,segura2013}; therefore, we
fit a simpler model in which we assumed density independence. Third,
observation error could bias parameter estimates \citep{knape2012} or mask our
ability to detect heavy tails \citep{ward2007}; therefore, we fit a model
where we allowed for a fixed quantity of observation error ($0.2$ standard
deviations on a log scale). Finally, the Gompertz model assumes that
population growth rate declines linearly with log abundance. We also fit an
alternative model, the Ricker-logistic model, which assumes that population
growth rate declines linearly with abundance itself (Supporting Information).

@@

We examined all time series with published explanations of why the black-swan
events occurred (Tables~\ref{tab:sparks} and \ref{tab:causes-supp}). The
majority of documented events (\percBSDown \%) were downward black swans and
involved a combination of multiple factors. For example, a synchronization of
environmental- and predation-mediated population cycles are thought to have
caused a downward black-swan event for a water vole (\textit{Arvicola
  terrestris}) population \citep{saucy1994}. Other black swans were the result
of a sequence of extreme climate events on their own. For instance, severe
winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan
downswings in grey heron (\textit{Ardea cinerea}) abundance in the United
Kingdom \citep{stafford1971}. Our analysis finds that the last event was a
combination of two black-swan events in a row and it took the population three
times longer to recover than predicted \citep{stafford1971}. Downwards black
swans were sometimes followed by upwards black swans. For example, during a
period of population crowding and nest shortages, a population of European
shag cormorants (\textit{Phalacrocorax aristotelis}) on the Farne Islands, UK,
declined suddenly following a red tide event in 1968 \citep{potts1980}. This
freed quality nest sites for first-time breeders, productivity rapidly
increased, and the population experienced a rapid upswing in abundance.
\citep{potts1980}.

@@

We found strong evidence for black swans (heavy-tailed process noise) in
\overallMinPerc--\overallMaxPerc\% of ecological time series. Black swans were
usually (\percBSDown \%) downward events and were detected more frequently in
longer time series and in populations with a smaller magnitude of process
noise. Black swans were not associated with density dependence, population
growth rate, or lifespan. In verified cases, black-swan events were often a
result of the synchrony of extreme climate, predators and parasite cycles, and
strong changes in human pressures. Our empirical results, sensitivity
analyses, and simulation tests suggest that estimating the tail shape of
process noise is a viable method of detecting black-swan population dynamics
and if anything will underestimate the probability of black-swan events. The
presence of black swans highlights the importance of developing management
strategies that detect quickly, respond to, and are robust to extremes in
population dynamics---particularly as the frequency and magnitude of climatic
extremes increase over the next century \citep{easterling2000,ipcc2012}.

@@

There are a number of caveats when considering the generality of our results.
The GPDD data represent a taxonomically and geographically biased sample of
populations---the longer time series we focus on are dominated by commercially
and recreationally important species and a disproportionate number of
populations are located in the United Kingdom. Although we would expect to find
qualitatively similar evidence for black swans in many other large taxonomic or
geographic samples of populations, the common forces driving those black swans
would likely differ (e.g.~severe winters in Table~\ref{tab:sparks}). In
addition to a possibly biased sample of populations, some black-swan detections
could just be recording mistakes, or conversely, some extreme observations may
have been discarded or altered if they were erroneously suspected of being
recording mistakes. Indeed, three of the populations our method initially
identified as heavy tailed turned out to be data-entry errors and were
discarded (Supporting Information). A final caveat is that the temporal scales
of observation and population dynamics vary considerably across populations in
the GPDD and this likely influences the detection of heavy tails. As an
example, if we make frequent observations relative to generation time (e.g.~for
many large-bodied mammals) we will average across generations and perhaps miss
black swans. Conversely, if we census populations infrequently relative to
generation time (e.g.~many insects in the GPDD) the recorded data may average
across extreme and less-extreme events and also dampen black-swan dynamics.

@@

Most importantly, what is the impact of allowing for black swans in forecasts
of ecological risk? In recent decades, ecology has moved toward focussing on
aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a,
  thompson2013}. Our results suggest that an added focus on ecological
extremes represents the next frontier, particularly in the face of increased
climate extremes \citep{meehl2004,thompson2013,ipcc2012}. Financial analysts
are concerned with the shape of the tails of financial returns because these
directly impact estimates of risk---the probability of a specific magnitude of
undesired event occurring \citep{rachev2008}. A comparable focus in ecology
would increase our estimates of extinction risk, since these would be
disproportionately impacted by downward black-swan events.

@@

\bibliographystyle{ecologyletters}
\bibliography{/Users/seananderson/Dropbox/tex/jshort,/Users/seananderson/Dropbox/tex/ref3}

>>>>

Black swans are statistically improbable events that nonetheless occur---often
with profound consequences. While extremes in the physical environment, such
as monsoons and heat waves, are widely studied and increasing in magnitude and
frequency, it remains unclear the extent to which ecological populations
buffer or suffer from such extremes. Here, estimate the degree of
heavy-tailedness (presence of black swans) in ecological process noise by
applying a probability model to \NPops~time series from around the world
across \NOrders~taxonomic orders and seven classes. We find strong evidence of
black swans, but they are rare, occurring in
\overallMinPerc--\overallMaxPerc\% of populations: most frequently for birds
(\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), and insects
(\InsectaRangePerc\%). Black swans were predominantly (\percBSDown \%)
downward events and were not explained by any life-history covariates, but
tended to be driven by external perturbations such as climate, severe winters,
predator or parasite cycles, and the combined effects of multiple factors.
Extreme events were more frequently detected for populations with longer time
series and lower levels of process noise; for shorter and noisier time series,
our simulations suggested black-swan dynamics are often not identifiable as
such. The presence of black swans in population dynamics highlights the
importance of developing robust conservation and management
strategies---particularly as the frequency and magnitude of climate extremes
increase over the next century.

@@

Black swans are unexpected extreme events with potentially dramatic
consequences \citep{taleb2007,sornette2009}. One of the most striking black
swans in ecology is the asteroid collision that may have marked the
end-Cretaceous mass extinction 65 million years ago
\citep{alvarez1980,harnik2012}. Today, climate extremes, in concert with
shifts in mean temperature, are expected to cause the greatest ecological and
societal damage \citep{ipcc2012}. But, while extremes in the physical
environment such as wave height, storm severity, and temperature are frequent
events \citep{gaines1993,katz2005}, it remains unclear the extent to which
ecological systems buffer or suffer from black swans \citep{nunez2012}.

There is compelling anecdotal evidence for ecological black swans, but
systematic evidence across taxa has been elusive. A survey of ecologists
indicated that surprising outcomes of field experiments are far more common
than we assume \citep{doak2008}, and events such as the global invasion of
Argentine ants and the mutation of viruses to infect new hosts could be
considered black swans \citep{nunez2012}. In fact, anecdotes of population
catastrophes are numerous and catastrophes may be the most important element
affecting population persistence \citep{mangel1994}. For marine mammal
populations, we have compelling evidence of catastrophes \citep{gerber2001,
  ward2007}, and recently, time-series of marine microbe abundance
\citep{segura2013} and time to extinction for experimental waterflea
populations \citep{drake2014} have been found to follow heavier tailed
distributions than the normal. Despite these examples, as far as we can tell
there have been two systematic surveys for ecological black swans: one with
North American breeding birds and the other with the Global Population
Dynamics Database, which uncovered little clear evidence for them
\citep{keitt1998,allen2001,halley2002}. Nevertheless there are methodological
challenges to their detection \citep{allen2001,ward2007}.

@@

Our main analysis focuses on the commonly applied Gompertz population dynamics
model \citep[e.g.][]{knape2012,dennis2014,connors2014}. The Gompertz model
represents population growth as a linear function in $\log$ space. If we let
$x_t$ represent the $\log$ abundance ($N$) at time $t$, we can represent the
Gompertz model as:
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma).
\end{align*}
The growth parameter $\lambda$ represents the expected growth rate if $N_t =
1$. The model is density independent if $b = 1$, maximally density dependent
if $b = 0$, and inversely density dependent if $b < 0$. Usually, the process
noise $\epsilon_t$ is modelled as normally distributed, but in our paper we
assume it is drawn from a t distribution with scale parameter $\sigma$ and
degrees of freedom $\nu$. In previous analyses of the GPDD, the Gompertz was
most often identified as the most parsimonious population model fit to these
data \citep{brook2006}.

@@

We chose weakly informative priors to incorporate our understanding of
plausible population dynamics (\citeauthor{gelman2014} \citeyear{gelman2014};
Figs~\ref{fig:didactic}f--h, \ref{fig:priors}). For $\nu$, we chose an
exponential prior with rate parameter of $0.01$ truncated at values above
two---a slightly less informative prior than suggested by
\citet{fernandez1998}. This prior gives only a \basePriorProbHeavy \%
probability that $\nu < 10$ but constrains the sampling sufficiently to avoid
wandering off towards infinity. In any case, for $\nu > 20$ the t distribution
is almost indistinguishable from the normal distribution
(Fig.~\ref{fig:didactic}). Based on the shape of the t distribution, we chose
the probability that $\nu < 10$, Pr($\nu < 10$), to define the probability of
heavy-tailed dynamics. When categorizing a population as heavy or normal tailed,
we used a threshold of 0.5 probability.

We fit alternative population models to test if four key phenomena
systematically changed our conclusions. Autocorrelation has been suggested as
a reason for increased observed variability of abundance time series through
time, which could create apparent heavy tails \citep{inchausti2002};
therefore, we fit a model that included serial correlation in the residuals.
Additionally, previous work has modelled abundance or growth rates without
accounting for density dependence \citep{halley2002,segura2013}; therefore, we
fit a simpler model in which we assumed density independence. Third,
observation error could bias parameter estimates \citep{knape2012} or mask our
ability to detect heavy tails \citep{ward2007}; therefore, we fit a model
where we allowed for a fixed quantity of observation error ($0.2$ standard
deviations on a log scale). Finally, the Gompertz model assumes that
population growth rate declines linearly with log abundance. Therefore, we
also fit an alternative model, the Ricker-logistic model, which assumes that
population growth rate declines linearly with abundance itself (Supporting
Information).

@@

We examined all time series with published explanations of why the black-swan
events occurred (Tables~\ref{tab:sparks} and \ref{tab:causes-supp}). The
majority of documented events (\percBSDown \%) were downward black swans and
involved a combination of multiple factors. For example, a synchronization of
environmental- and predation-mediated population cycles are thought to have
caused a downward black-swan event for a water vole (\textit{Arvicola
  terrestris}) population \citep{saucy1994}. Other black swans were the result
of a sequence of extreme climate events on their own. For instance, severe
winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan
downswings in grey heron (\textit{Ardea cinerea}) abundance in the United
Kingdom \citep{stafford1971}. Our analysis finds that the last event was a
combination of two black-swan events in a row and it took the population three
times longer to recover than predicted \citep{stafford1971}. Downwards black
swans were sometimes followed by upwards black swans. For example, during a
period of population crowding and nest shortages, a population of European
shag cormorants (\textit{Phalacrocorax aristotelis}) on the Farne Islands,
United Kingdom, declined suddenly following a red tide event in 1968
\citep{potts1980}. This freed quality nest sites for first-time breeders,
productivity rapidly increased, and the population experienced a rapid upswing
in abundance. \citep{potts1980}.

@@

We found strong evidence for black swans (heavy-tailed process noise) in
\overallMinPerc--\overallMaxPerc\% of ecological time series. Black swans were
usually (\percBSDown \%) downward events and were detected more frequently in
longer time series and in populations with a smaller magnitude of process
noise. Black swans were not associated with density dependence, population
growth rate, or lifespan. In verified cases, black-swan events were often a
result of the combination of extreme climate, predators and parasite cycles,
and strong changes in human pressures. Our empirical results, sensitivity
analyses, and simulation tests suggest that estimating the tail shape of
process noise is a viable method of detecting black-swan population dynamics
and if anything will underestimate the probability of black-swan events. The
presence of black swans highlights the importance of developing management
strategies that detect quickly, respond to, and are robust to extremes in
population dynamics---particularly as the frequency and magnitude of climatic
extremes increase over the next century \citep{easterling2000,ipcc2012}.

@@

There are a number of caveats when considering the generality of our results.
The GPDD data represent a taxonomically and geographically biased sample of
populations---the longer time series we focus on are dominated by commercially
and recreationally important species and a disproportionate number of
populations are located in the United Kingdom. Although we would expect to
find qualitatively similar evidence for black swans in many other large
taxonomic or geographic samples of populations, the common forces driving
those black swans would likely differ (e.g.~severe winters in
Table~\ref{tab:sparks}). In addition to a possibly biased sample of
populations, some black-swan detections could just be recording mistakes, or
conversely, some extreme observations may have been discarded or altered if
they were erroneously suspected of being recording mistakes. Indeed, we
discarded three of the populations that our method initially identified as
heavy tailed because they turned out to be data-entry errors (Supporting
Information). A final caveat is that the temporal scales of observation and
population dynamics vary considerably across populations in the GPDD and this
likely influences the detection of heavy tails. As an example, if we make
frequent observations relative to generation time (e.g.~for many large-bodied
mammals) we will average across generations and perhaps miss black swans.
Conversely, if we census populations infrequently relative to generation time
(e.g.~many insects in the GPDD) the recorded data may average across extreme
and less-extreme events and also dampen black-swan dynamics.

@@

Most importantly, what is the impact of allowing for black swans in forecasts
of ecological risk? Even extremely rare catastrophes can have a profound
influence on population persistence \citep{mangel1994}. In recent decades,
ecology has moved toward focussing on aspects of variance in addition to mean
responses \citep[e.g.][]{loreau2010a, thompson2013}. Our results suggest that
an added focus on ecological extremes represents the next frontier,
particularly in the face of increased climate extremes \citep{meehl2004,
  thompson2013, ipcc2012}. Financial analysts are concerned with the shape of
the downward tails of financial returns because these directly impact
estimates of risk---the probability of a specific magnitude of undesired event
occurring \citep{rachev2008}. A comparable focus in ecology would increase our
estimates of extinction risk, since these would be disproportionately impacted
by downward black-swan events.

@@

\bibliographystyle{apalike}
\bibliography{/Users/seananderson/Dropbox/tex/jshort,/Users/seananderson/Dropbox/tex/ref3}
